Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models

### 2025-02-14 brms_meeg.qmd
pushin new simulation results and updated title and abstract

---
title: "Modelling M/EEG data with Bayesian generalised additive multilevel models"
shorttitle: "Bayesian M/EEG modelling"
author:
  - name: Ladislas Nalborczyk
    corresponding: true
    orcid: 0000-0002-7419-9855
    email: ladislas.nalborczyk@cnrs.fr
    url: https://lnalborczyk.github.io
    affiliations:
      - id: id1
        name: "Aix Marseille Univ, CNRS, LPL"
        address: "5 avenue Pasteur"
        city: "13100 Aix-en-Provence"
        country: France
  - name: Paul Bürkner
    orcid: 0000-0001-5765-8995
    url: https://paulbuerkner.com
    affiliations:
      - name: "TU Dortmund University, Department of Statistics"
author-note:
  disclosures:
    conflict-of-interest: "The authors have no conflicts of interest to disclose."
    # financial-support: "This study was supported by Grant [Grant Number] from [Funding Source]."
abstract: "Time-resolved electrophysiological measurements such as those offered by magneto- or electro-encephalography (M/EEG) provide a unique window onto neural activity underlying cognitive process and how they unfold over time. Typically, we are interested in testing whether such measures differ across conditions and/or groups. The conventional approach consists in conducting mass-univariate statistics through followed by some form of multiplicity correction (e.g., FDR, FWER) or cluster-based inference. However, these cluster-based methods have an important downside: they shift the focus of inference from the timepoint to the cluster level, thus preventing any conclusion to be made about the onset and offset of effects (e.g., differences across conditions). Here, we introduce a novel *model-based approch* for analysing one-dimensional M/EEG timeseries such as ERPs or decoding timecourses and their differences across conditions or group. This approach relies on Bayesian generalised additive multilevel models, which output the posterior probabilility of the effect being above 0 (or above chance) at every timestep, while naturally taking into account the temporal dependencies and between-subject variability present in such data."
keywords: [EEG, MEG, generalised additive models, mixed-effects models, multilevel models, Bayesian statistics, brms]
floatsintext: true
numbered-lines: true
bibliography:
  - meeg_modelling.bib
  - grateful-refs.bib
suppress-title-page: false
link-citations: true
# see https://www.overleaf.com/learn/latex/Font_typefaces
# and systemfonts::system_fonts()
# pdf-engine: xelatex
# monofont: Roboto
mainfont: CMU Serif
fontsize: 10pt
draftfirst: false
draftall: false
a4paper: true
lang: en
toc: true
language:
  citation-last-author-separator: "&"
  email: "email"
  title-block-author-note: "Author note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
format:
  # apaquarto-html: default
  apaquarto-pdf:
    latex-engine: xelatex
    documentmode: doc # man, jou, doc, stu
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{mathtools}
---

@@

# template for citing R packages
pkrt_set(pkg = "the `R` package `:pkg` v :ver [:ref]")
```

@@

Here are some useful references to be discussed [@combrisson_exceeding_2015; @ehinger_unfold_2019; @hayasaka_validating_2003; @luck_how_2017; @pedersen_hierarchical_2019; @frossard2021; @frossard2022; @gramfort2013; @maris2007; @pernet2015]... See also [@maris2011]... and [@rosenblatt2018] (history of cluster-based approaches and using a data split?)... Cluster failure [@eklund2016]...

In the following, we consider two approaches to modelling the non-linear timecourse of M/EEG ERPs (or decoding performance): i) generalised additive models (GAMs) with smoothing splines [@wood2003; @wood2004]...

and ii) Gaussian processes (GPs) with a smooth covariance kernel [@rasmussen2005] and low-rank approximation [@riutort-mayol_practical_2023].

## Previous modelling work

Recent example of GLM for EEG [@wüllhorst2025; @fischer2013]... "Similar approaches have been successfully applied to EEG time- (Rousselet et al., 2008) and frequency-domain (Cohen and Cavanagh, 2011) data and allow the simultaneous investigation of multiple independent variables while preserving the high temporal resolution of the EEG."... See also [@rousselet2008; @hauk2006]... Example of two-stage regression analysis [i.e., individual-level then group-level, @dunagan2024]...

From @dimigen2021: "Recently, spline regression has been applied to ERPs (Hendrix, Baayen, & Bolger, 2017; Kryuchkova,Tucker, Wurm, & Baayen,2012; Tremblay & Baayen, 2010; Tremblay & Newman, 2015)"... GAMMs for EEG data [@abugaber2023; @meulman2015]...

Disentangling overlapping processes [@skukies_brain_2024; @skukies_modelling_2021]... Weighting single trials [@pernet2022]... The LIMO toolbox (https://onlinelibrary.wiley.com/doi/10.1155/2011/831409)... Using Bayes factors [@teichmann2022]...

@@

See for instance these tutorials [@sóskuthy2017; @winter2016] or application to phonetic data [@wieling2018] or this introduction [@baayen2020] or these reference books [@hastie2017; @wood2017]...

@@

$$
\begin{aligned} 
y_{i} &\sim \mathrm{EF}\left(\mu_{i}, \phi\right)\\
g\left(\mu_i\right) &= A_{i} + \mathbf{X}_{i} \gamma + \sum_{j=1}^{J} f_{j}\left(x_{ij}\right)
\end{aligned}
$$

@@

where $\beta_{jk}$ is the weight (coefficient) associated with the $k$th basis function $b_{jk}()$ evaluated at the covariate value $x_{ij}$ for the $j$th smooth function $f_{j}$. Splines' coefficients are penalised...

@@

```{r fig-eeg, echo = FALSE, fig.width = 9, fig.cap = "Some ERPs in two conditions with 50 trials each, for a group of 20 participants."}
# importing R version of Matlab code from Yeung et al. (2004)
source("code/eeg_noise.R")

@@

# plotting the data
raw_df %>%
    summarise(eeg = mean(eeg), .by = c(participant, condition, time) ) %>%
    ggplot(aes(x = time, y = eeg, colour = condition) ) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_vline(xintercept = true_offset, linetype = 2) +
    geom_line(show.legend = FALSE) +
    facet_wrap(~participant) +
    scale_colour_met_d(name = "Johnson") +
    labs(x = "Time (s)", y = "EEG signal (a.u.)")
```

@@

```{r fig-erp, echo = FALSE, out.width = "75%", fig.cap = "Group-level average difference between conditions (mean +/- standard error of the mean). The 'true' onset and offset are indicated by the vertical dashed lines."}
# averaging across participants
raw_df %>%
    pivot_wider(names_from = condition, values_from = eeg, values_fn = mean) %>%
    mutate(eeg_diff = cond2 - cond1) %>%
    summarise(
        eeg_mean = mean(eeg_diff),
        eeg_sem = sd(eeg_diff) / sqrt(n() ),
        .by = time
        ) %>%
    mutate(lower = eeg_mean - eeg_sem, upper = eeg_mean + eeg_sem) %>%
    # plotting the data
    ggplot(aes(x = time, y = eeg_mean) ) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_vline(xintercept = true_offset, linetype = 2) +
    geom_ribbon(
        aes(ymin = lower, ymax = upper, colour = NULL),
        alpha = 0.25, show.legend = FALSE
        ) +
    geom_line(show.legend = FALSE) +
    labs(x = "Time (s)", y = "EEG difference (a.u.)")
```

@@

We then fitted a Bayesian GAM using the `brms` package [@brms2017; @brms2018; @nalborczyk2019]. We used the default priors in `brms`, that is, weakly informative priors. We ran eight Markov Chain Monte-Carlo (MCMC) to approximate the posterior distribution, including each 5000 iterations and a warmup of 2000 iterations, yielding a total of $8 \times (5000-2000) = 24000$ posterior samples to use for inference. Posterior convergence was assessed examining trace plots as well as the Gelman–Rubin statistic $\hat{R}$.

@@

@@

    # plotting it
    prob_y_above_0 %>%
        ggplot(aes(x = time, y = m) ) +
        geom_hline(yintercept = 0, linetype = 2) +
        geom_vline(xintercept = true_onset, linetype = 2) +
        geom_line() +
        geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
        scale_colour_met_d(name = "Johnson") +
        scale_fill_met_d(name = "Johnson") +
        labs(
            x = "Time (s)",
            y = expression(p(beta~"|"~data, model) )
            )

@@

```{r fig-post-prob-test, echo = FALSE, out.width = "75%", fig.asp = 0.5, fig.cap = "Posterior probability of the ERP difference (slope) being above 0 according to the GAM."}
# defining a function to plot posterior prob of slope above 0
plot_post_test <- function (model, data = NULL, eps = 0.1, n_approx = 1e3, n_draws = 1e3) {

@@

```{r fig-post-prob-ratio, echo = FALSE, fig.width = 8, fig.asp = 0.75, fig.cap = "Ratio of posterior probability according to the GAM. Timesteps above threshold (10) are highlighted in green."}
# defining some parameters
n_approx <- 1e3
eps <- 0.05
threshold <- 10

@@

We compared the ability of the GAMM to correctly estimate the onset and offset of the ERP difference to widely used methods. First, we conducted mass-univariate t-tests (thus treating each timestep independently) and identified the onset and offset of the ERP difference as the first and last values crossing an arbitrary significance threshold ($\alpha = 0.01$). We then followed the same approach but after applying different forms of multiplicity correction the the p-values. We compared two methods that control the false discovery rate (FDR) [i.e., BH95, @benjamini1995; and BY01, @benjamini2001], one method that controls the familywise error rate (FWER) (i.e., Holm–Bonferroni method, https://www.jstor.org/stable/4615733), and two cluster-based methods [permutation with a single threshold and TFCE, @smith2009]. The BH95, BY01, and Holm corrections were applied to the p-values using the `p.adjust()` function in `R`. The cluster-based inference was implemented using a cluster-sum statistic of squared t-values, as implemented in `MNE-Python` [@gramfort2013], called via `r pkrt("reticulate")`. We also compared these estimates to the onset and offset as estimated using the binary segmentation algorithm, as implemented in `r pkrt("changepoint")`, and applied directly to the squared t-values [as in @rousselet_using_2025]. For visualisation and interpretability purposes, we converted p-values to s-values, which can be interpreted as bits of surprising information, assuming a null effect [@greenland2019] (@fig-corrections).

@@

Onset/offset estimation methods were assessed using these summary statistics of the Monte Carlo sampling distributions of onsets: median absolute error (MAE) and variance, with 10.000 iterations (i.e., 10.000 simulated datasets)...

@@

-   Create many (e.g., 10 or 20) half train/test splits of the data
-   For each fold, estimate the onset/offset on both splits using all methods
-   Then summarise the distribution of between-split difference across folds

This will allow checking that the proposed approach produces more reliable onset/offset estimates.

@@

# computing and plotting bias, MAE, and variance
# bias := true_onset - median(sim_onset_distrib)
# MAE :=  mean(abs(sim_onset-true_onset))
# variance := variance(sim_onset_distrib)
sim_results %>%
    # keeping only the best brms results
    filter(eps == 0 & threshold == 20) %>%
    select(-eps,-threshold) %>%
    pivot_longer(cols = -simulation_id, names_to = "measure", values_to = "value") %>%
    data.frame() %>%
    mutate(
        error = case_when(
            grepl("onset", measure) ~ abs(value - true_onset),
            grepl("offset", measure) ~ abs(value - true_offset)
            )
        ) %>%
    separate(measure, into = c("type", "method"), sep = "_", extra = "merge") %>%
    group_by(type, method) %>%
    summarise(
        MAE = median(error, na.rm = TRUE),
        variance = var(error, na.rm = TRUE),
        .groups = "drop"
        ) %>%
    ungroup() %>%
    mutate(type = factor(x = type, levels = c("onset", "offset") ) ) %>%
    mutate(
        method = factor(
            x = method,
            levels = c(
                "p", "bh", "by", "holm", "cluster_mass",
                "cluster_tfce", "cpt", "brms"
                ),
            labels = c(
                "p-value", "FDR BH95", "FDR BY01", "Holm", "Cluster-mass",
                "TFCE", "Change point", "brms"
                ),
            )
        ) %>%
    ggplot(
        aes(
            x = MAE,
            y = variance,
            colour = method,
            fill = method
            )
        ) +
    geom_point(
        size = 2,
        pch = 21,
        colour = "white",
        show.legend = FALSE
        ) +
    geom_label_repel(
        aes(label = method),
        colour = "white",
        size = 3,
        segment.color = NA,
        show.legend = FALSE
        ) +
    scale_y_log10() +
    facet_wrap(~type) +
    scale_fill_manual(values = met.brewer(name = "Johnson", n = 8) ) +
    scale_colour_manual(values = met.brewer(name = "Johnson", n = 8) ) +
    labs(x = "Median absolute error (s)", y = "Variance (log-scale)")
```

```{r fig-simulation-distribution, echo = FALSE, fig.width = 9, fig.asp = 0.5, fig.cap = "Distributions of onset and offset estimates for each method."}
sim_results %>%
    # keeping only the best brms results
    filter(eps == 0 & threshold == 20) %>%
    select(-eps,-threshold) %>%
    pivot_longer(
        cols = -simulation_id, 
        names_to = "measure",
        values_to = "value"
        ) %>%
    separate(measure, into = c("type", "method"), sep = "_", extra = "merge") %>%
    data.frame() %>%
    mutate(type = factor(x = type, levels = c("onset", "offset") ) ) %>%
    mutate(
        method = factor(
            x = method,
            levels = c(
                "p", "bh", "by", "holm", "cluster_mass",
                "cluster_tfce", "cpt", "brms"
                ),
            labels = c(
                "p-value", "FDR BH95", "FDR BY01", "Holm",
                "Cluster-mass", "TFCE", "Change point", "brms"
                ),
            )
        ) %>%
    mutate(
        plot_row = ifelse(
            method %in% c("p-value", "FDR BH95", "FDR BY01", "Holm"), 0, 1
            )
        ) %>%
    ggplot(
        aes(
            x = value,
            colour = method,
            fill = method
            )
        ) +
    geom_vline(xintercept = true_onset) +
    geom_vline(xintercept = true_offset) +
    geom_density(aes(fill = NULL) ) +
    # facet_wrap(~type, scales = "free_x") +
    # facet_wrap(~type) +
    facet_grid(plot_row~type, scales = "free_x") +
    theme_light(base_size = 12, base_family = "Open Sans") +
    scale_fill_manual(values = met.brewer(name = "Johnson", n = 8) ) +
    scale_colour_manual(values = met.brewer(name = "Johnson", n = 8) ) +
    labs(x = "Onset/Offset (s)", y = "Density")
```

@@

@@

# Using the `R` package and integration with `MNE-Python`

Explain how to use the `R` package and to integrate it with `MNE` epochs...

```{r r-python, eval = FALSE}
# to-do adding some code here...
```

>>>>

---
title: "Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models"
shorttitle: "Bayesian modelling of M/EEG data"
author:
  - name: Ladislas Nalborczyk
    corresponding: true
    orcid: 0000-0002-7419-9855
    email: ladislas.nalborczyk@cnrs.fr
    url: https://lnalborczyk.github.io
    affiliations:
      - id: id1
        name: "Aix Marseille Univ, CNRS, LPL"
        address: "5 avenue Pasteur"
        city: "13100 Aix-en-Provence"
        country: France
  - name: Paul Bürkner
    orcid: 0000-0001-5765-8995
    url: https://paulbuerkner.com
    affiliations:
      - name: "TU Dortmund University, Department of Statistics"
author-note:
  disclosures:
    conflict-of-interest: "The authors have no conflicts of interest to disclose."
    # financial-support: "This study was supported by Grant [Grant Number] from [Funding Source]."
abstract: "Time-resolved electrophysiological measurements such as those offered by magneto- or electro-encephalography (M/EEG) provide a unique window onto neural activity underlying cognitive processes. Typically, researchers are interested in testing whether and when such measures differ across conditions and/or groups. The conventional approach consists in conducting mass-univariate statistics through time followed by some form of multiplicity correction (e.g., FDR, FWER) or cluster-based inference. However, these cluster-based methods have an important downside: they shift the focus of inference from the timepoint to the cluster level, thus preventing any conclusion to be made about the onset or offset of effects (e.g., differences across conditions or groups). Here, we introduce a *model-based* approch for analysing M/EEG timeseries such as ERPs or timecourses of decoding performance and their differences across conditions or groups. This approach relies on Bayesian generalised additive multilevel models, which output the posterior probabilility of the effect being above 0 (or above chance) at every timestep, while naturally taking into account the temporal dependencies and between-subject variability present in such data. Using both simulation and actual EEG data, we show that the proposed approach largely outperforms conventional methods in determining both the onset and offset of M/EEG effects (e.g., ERPs difference, decoding performance), producting more precise and more reliable estimates. We provide an R package implementing the approach and illustrate how to integrate it into M/EEG statistical pipelines in MNE-Python."
keywords: [EEG, MEG, generalised additive models, mixed-effects models, multilevel models, Bayesian statistics, brms]
floatsintext: true
numbered-lines: true
bibliography:
  - meeg_modelling.bib
  - grateful-refs.bib
suppress-title-page: false
link-citations: true
# see https://www.overleaf.com/learn/latex/Font_typefaces
# and systemfonts::system_fonts()
# pdf-engine: xelatex
# monofont: Roboto
mainfont: CMU Serif
fontsize: 10pt
draftfirst: false
draftall: false
a4paper: true
donotrepeattitle: true
lang: en
toc: true
language:
  citation-last-author-separator: "&"
  email: "email"
  title-block-author-note: "Author note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
format:
  # apaquarto-html: default
  apaquarto-pdf:
    latex-engine: xelatex
    documentmode: doc # man, jou, doc, stu
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{mathtools}
---

@@

# template for citing R packages
# pkrt_set(pkg = "the `R` package `:pkg` [:ref]")
pkrt_set(pkg = "the `R` package `:pkg` v :ver [:ref]")
```

@@

Here are some useful references to be discussed [@combrisson_exceeding_2015; @ehinger_unfold_2019; @hayasaka_validating_2003; @luck_how_2017; @pedersen_hierarchical_2019; @frossard2021; @frossard2022; @gramfort2013; @maris2007; @pernet2015]... See also @maris2011 and @rosenblatt2018 (history of cluster-based approaches)... Clusters failures [@eklund2016]...

## Previous work

Recent example of GLM for EEG [@wüllhorst2025; @fischer2013]... See also [@rousselet2008; @hauk2006]... Example of two-stage regression analysis [i.e., individual-level then group-level, @dunagan2024]...

From @dimigen2021: Recently, spline regression has been applied to ERPs [e.g., @hendrix2017; @kryuchkova2012]... GAMMs for EEG data [@abugaber2023; @meulman2015]...

Disentangling overlapping processes [@skukies_brain_2024; @skukies_modelling_2021]... Weighting single trials [@pernet2022]... The LIMO toolbox [@pernet2011]...

Using Bayes factors for decoding performance [@teichmann2022]...

## Bayesian regression modelling

Short intro/recap about Bayesian (linear and generalised) regression models...

@@

See for instance these tutorials [@sóskuthy2017; @winter2016] or application to phonetic data [@wieling2018] or this introduction [@baayen2020] or these reference books [@hastie2017; @wood2017]... application to pupillometry [@vanrij2019]... GAMLSS for neuroimaging data [@dinga2021]...

@@

$$
\begin{aligned} 
y_{i} &\sim \mathrm{EF}\left(\mu_{i}, \phi\right)\\
g\left(\mu_i\right) &=  \underbrace{A_{i} + \mathbf{X}_{i} \gamma}_{\text{parametric part}} + \underbrace{\sum_{j=1}^{J} f_{j}\left(x_{ij}\right)}_{\text{non-parametric part}}
\end{aligned}
$$

@@

where $\beta_{jk}$ is the weight (coefficient) associated with the $k$th basis function $b_{jk}()$ evaluated at the covariate value $x_{ij}$ for the $j$th smooth function $f_{j}$. Splines coefficients are penalised (usually through the squared of the smooth functions' second derivative) in a way that can be interpreted as a prior on the "wiggliness" of the function...

## Bayesian generalised additive multilevel models

Now describe the Bayesian GAMM... Proper inclusion of varying/random effects in the model specification protects against overly wiggly curves [@baayen2020]...

@@

```{r fig-eeg, echo = FALSE, fig.width = 9, fig.cap = "Averaged (mean) simulated EEG activity in two conditions with 50 trials each, for a group of 20 participants. The error band represents the mean plus/minus 1 standard error of the mean."}
# importing R version of Matlab code from Yeung et al. (2004)
source("code/eeg_noise.R")

@@

# plotting the data
raw_df %>%
    summarise(
        eeg_mean = mean(eeg),
        eeg_se = sd(eeg) / sqrt(n() ),
        .by = c(participant, condition, time),
        ) %>%
    ggplot(aes(x = time, y = eeg_mean, colour = condition, fill = condition) ) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_vline(xintercept = true_offset, linetype = 2) +
    geom_ribbon(
        aes(ymin = eeg_mean - eeg_se, ymax = eeg_mean + eeg_se, colour = NULL),
        alpha = 0.3, show.legend = FALSE
        ) +
    geom_line(show.legend = FALSE) +
    facet_wrap(~participant) +
    scale_colour_met_d(name = "Johnson") +
    labs(x = "Time (s)", y = "EEG signal (a.u.)")
```

@@

```{r fig-erp, echo = FALSE, out.width = "75%", fig.cap = "Group-level average difference between conditions (mean +/- standard error of the mean). The 'true' onset and offset are indicated by the vertical dashed lines."}
# averaging across participants
raw_df %>%
    pivot_wider(names_from = condition, values_from = eeg, values_fn = mean) %>%
    mutate(eeg_diff = cond2 - cond1) %>%
    summarise(
        eeg_mean = mean(eeg_diff),
        eeg_sem = sd(eeg_diff) / sqrt(n() ),
        .by = time
        ) %>%
    mutate(lower = eeg_mean - eeg_sem, upper = eeg_mean + eeg_sem) %>%
    # plotting the data
    ggplot(aes(x = time, y = eeg_mean) ) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_vline(xintercept = true_onset, linetype = 2) +
    geom_vline(xintercept = true_offset, linetype = 2) +
    geom_ribbon(
        aes(ymin = lower, ymax = upper, colour = NULL),
        alpha = 0.3, show.legend = FALSE
        ) +
    geom_line(show.legend = FALSE) +
    labs(x = "Time (s)", y = "EEG difference (a.u.)")
```

@@

We then fitted a Bayesian GAM using the `brms` package [@brms2017; @brms2018; @nalborczyk2019]. We used the default priors in `brms`, that is, weakly informative priors. We ran eight Markov Chain Monte-Carlo (MCMC) to approximate the posterior distribution, including each 5000 iterations and a warmup of 2000 iterations, yielding a total of $8 \times (5000-2000) = 24000$ posterior samples to use for inference. Posterior convergence was assessed examining trace plots as well as the Gelman–Rubin statistic $\hat{R}$. The `brms` package uses the same syntax as `r pkrt("mgcv")` for specifying smooth effects.

@@

```{r gamm, echo = FALSE, eval = FALSE, message = FALSE}
# defining a contrast for condition
contrasts(raw_df$condition) <- c(-0.5, 0.5)

# fitting the GAMM with factor smooth (per participant)
gamm <- brm(
    # cubic regression splines with k-1 basis functions
    eeg ~ condition +
        s(time, bs = "cr", k = 10, by = condition) +
        s(time, participant, bs = "fs"),
    data = raw_df,
    family = gaussian(),
    warmup = 2000,
    iter = 5000,
    chains = 4,
    cores = 4,
    file = "models/gamm.rds"
    )
```

@@

    # plotting it
    prob_y_above_0 %>%
        ggplot(aes(x = time, y = m) ) +
        geom_hline(yintercept = 0, linetype = 2) +
        geom_vline(xintercept = true_onset, linetype = 2) +
        geom_vline(xintercept = true_offset, linetype = 2) +
        geom_line() +
        geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
        scale_colour_met_d(name = "Johnson") +
        scale_fill_met_d(name = "Johnson") +
        labs(
            x = "Time (s)",
            y = expression(p(beta~"|"~data, model) )
            )

@@

```{r fig-post-prob-test, echo = FALSE, out.width = "75%", fig.asp = 0.75, fig.cap = "Posterior probability of the ERP difference (slope) being above 0 according to the GAM."}
# defining a function to plot posterior prob of slope above 0
plot_post_test <- function (model, data = NULL, eps = 0.1, n_approx = 1e3, n_draws = 1e3) {

@@

```{r fig-post-prob-ratio, echo = FALSE, out.width = "75%", fig.asp = 0.75, fig.cap = "Ratio of posterior probability according to the GAM. Timesteps above threshold (10) are highlighted in green."}
# defining some parameters
n_approx <- 1e3
eps <- 0.05
threshold <- 10

@@

We compared the ability of the GAMM to correctly estimate the onset and offset of the ERP difference to widely used methods. First, we conducted mass-univariate t-tests (thus treating each timestep independently) and identified the onset and offset of the ERP difference as the first and last values crossing an arbitrary significance threshold ($\alpha = 0.01$). We then followed the same approach but after applying different forms of multiplicity correction the the p-values. We compared two methods that control the false discovery rate (FDR) [i.e., BH95, @benjamini1995; and BY01, @benjamini2001], one method that controls the familywise error rate (FWER) [i.e., Holm–Bonferroni method, @holm1979], and two cluster-based methods [permutation with a single threshold and TFCE, @smith2009]. The BH95, BY01, and Holm corrections were applied to the p-values using the `p.adjust()` function in `R`. The cluster-based inference was implemented using a cluster-sum statistic of squared t-values, as implemented in `MNE-Python` [@gramfort2013], called via `r pkrt("reticulate")`. We also compared these estimates to the onset and offset as estimated using the binary segmentation algorithm, as implemented in `r pkrt("changepoint")`, and applied directly to the squared t-values [as in @rousselet_using_2025]. For visualisation and interpretability purposes, we converted p-values to s-values, which can be interpreted as bits of surprising information, assuming a null effect [@greenland2019] (@fig-corrections).

@@

Onset/offset estimation methods were assessed using the median absolute error (MAE) and variance of 10.000 simulated datasets...

@@

-   Create many (e.g., 10 or 20) half train/test splits of the data
-   For each fold, estimate the onset/offset on both splits using all methods
-   Then summarise the distribution of onset/offset with the median and variance across folds

This will allow checking that the proposed approach produces reliable onset/offset estimates.

```{r reliability, echo = FALSE, eval = FALSE}
# library(caret)
# see https://github.com/topepo/caret/
# https://topepo.github.io/caret/data-splitting.html
# and createDataPartition()
# and createFolds()
# trainIndex <- createDataPartition(
#     y = summary_df$participant,
#     p = 0.5,
#     list = FALSE,
#     times = 10
#     )

# subsetting the data
# summary_df_train <- summary_df[trainIndex, ]
# summary_df_test  <- summary_df[-trainIndex, ]

# getting unique participants
unique_participants <- unique(summary_df$participant)

# getting the number of participants
num_participants <- length(unique_participants)

# ensuring participants are randomly shuffled
shuffled_participants <- sample(unique_participants)

# defining the number of splits
num_splits <- 10

# creating a dataframe where each participant appears in every split
split_df <- crossing(participant = unique_participants, split_number = 1:10)

# assigning alternating groups (A or B) within each split for each participant
# split_df <- split_df %>% mutate(group = rep(c("A", "B"), length.out = n() ) )
split_df <- split_df %>%
    group_by(split_number) %>%
    # shuffling groups for each split
    mutate(group = sample(rep(c("A", "B"), length.out = n() ) ) ) %>%
    ungroup()

# sanity checks
# table(split_df$group)
# split_df %>%
#     ggplot(aes(x = group) ) +
#     geom_histogram(stat = "count") +
#     facet_wrap(~participant, ncol = 5)

# now for each split, assess the onset and offset
participants_split1_train <- split_df %>%
    filter(split_number == 1 & group == "A") %>%
    pull(participant)
participants_split1_test <- split_df %>%
    filter(split_number == 1 & group == "B") %>%
    pull(participant)

# split1 data
summary_df_split1_train <- summary_df %>%
    filter(participant %in% participants_split1_train)

# fitting the GAM
meta_gam_split1_train <- brm(
    eeg_mean | se(eeg_sd) ~
        condition + s(time, bs = "cr", k = 10, by = condition) +
        (1 | participant),
    data = summary_df_split1_train,
    family = gaussian(),
    warmup = 2000,
    iter = 5000,
    chains = 4,
    cores = 4
    )
```

@@

# computing and plotting bias, MAE, and variance
# bias := true_onset - median(sim_onset_distrib)
# MAE :=  mean(abs(sim_onset-true_onset))
# variance := variance(sim_onset_distrib)
sim_results %>%
    # keeping only the best brms results
    filter(eps == 0 & threshold == 20) %>%
    select(-eps,-threshold) %>%
    # filtering some alpha_level
    filter(alpha_level == 0.05) %>%
    select(-alpha_level) %>%
    pivot_longer(cols = -simulation_id, names_to = "measure", values_to = "value") %>%
    data.frame() %>%
    mutate(
        error = case_when(
            grepl("onset", measure) ~ abs(value - true_onset),
            grepl("offset", measure) ~ abs(value - true_offset)
            )
        ) %>%
    separate(measure, into = c("type", "method"), sep = "_", extra = "merge") %>%
    group_by(type, method) %>%
    summarise(
        MAE = median(error, na.rm = TRUE),
        variance = var(error, na.rm = TRUE),
        .groups = "drop"
        ) %>%
    ungroup() %>%
    mutate(type = factor(x = type, levels = c("onset", "offset") ) ) %>%
    mutate(
        method = factor(
            x = method,
            levels = c(
                "p", "bh", "by", "holm", "cluster_mass",
                "cluster_tfce", "cpt", "brms"
                ),
            labels = c(
                "Raw p-value", "FDR BH95", "FDR BY01", "Holm", "Cluster-mass",
                "TFCE", "Change point", "brms"
                ),
            )
        ) %>%
    ggplot(
        aes(
            x = MAE,
            y = variance,
            colour = method,
            fill = method
            )
        ) +
    geom_point(
        size = 2,
        pch = 21,
        colour = "white",
        show.legend = FALSE
        ) +
    geom_label_repel(
        aes(label = method),
        colour = "white",
        size = 3,
        segment.color = NA,
        show.legend = FALSE
        ) +
    scale_y_log10() +
    facet_wrap(~type) +
    scale_fill_manual(values = met.brewer(name = "Johnson", n = 8) ) +
    scale_colour_manual(values = met.brewer(name = "Johnson", n = 8) ) +
    labs(x = "Median absolute error (s)", y = "Variance (log-scale)")
```

```{r fig-simulation-distribution, echo = FALSE, fig.width = 9, fig.asp = 0.5, fig.cap = "Distributions of onset and offset estimates for each method."}
sim_results %>%
    # keeping only the best brms results
    filter(eps == 0 & threshold == 20) %>%
    # filtering some alpha_level
    filter(alpha_level == 0.05) %>%
    select(-alpha_level) %>%
    select(-eps,-threshold) %>%
    pivot_longer(
        cols = -simulation_id, 
        names_to = "measure",
        values_to = "value"
        ) %>%
    separate(measure, into = c("type", "method"), sep = "_", extra = "merge") %>%
    data.frame() %>%
    mutate(type = factor(x = type, levels = c("onset", "offset") ) ) %>%
    mutate(
        method = factor(
            x = method,
            levels = c(
                "p", "bh", "by", "holm", "cluster_mass",
                "cluster_tfce", "cpt", "brms"
                ),
            labels = c(
                "Raw p-value", "FDR BH95", "FDR BY01", "Holm",
                "Cluster-mass", "TFCE", "Change point", "brms"
                ),
            )
        ) %>%
    mutate(
        plot_row = ifelse(
            method %in% c("p-value", "FDR BH95", "FDR BY01", "Holm"), 0, 1
            )
        ) %>%
    ggplot(
        aes(
            x = value,
            colour = method,
            fill = method
            )
        ) +
    geom_vline(xintercept = true_onset) +
    geom_vline(xintercept = true_offset) +
    geom_density(aes(fill = NULL) ) +
    facet_grid(plot_row~type, scales = "free_x") +
    theme_light(base_size = 12, base_family = "Open Sans") +
    scale_fill_manual(values = met.brewer(name = "Johnson", n = 8) ) +
    scale_colour_manual(values = met.brewer(name = "Johnson", n = 8) ) +
    labs(x = "Onset/Offset (s)", y = "Density")
```

@@

We kept the exemplary models simple, but can be extended by adding varying/random effects (intercept and slope) for item (e.g., word)... but also continuous predictors at the trial level?

@@

# Integration with `MNE-Python`

Explain how to use the `R` package with `MNE` epochs...

```{r r-python, eval = FALSE}
# suggest a package name
library(available)
suggest(
    text = "Time-resolved electrophysiological measurements such as those offered by magneto- or electro-encephalography (M/EEG) provide a unique window onto neural activity underlying cognitive processes. Typically, researchers are interested in testing whether and when such measures differ across conditions and/or groups. The conventional approach consists in conducting mass-univariate statistics through time followed by some form of multiplicity correction (e.g., FDR, FWER) or cluster-based inference. However, these cluster-based methods have an important downside: they shift the focus of inference from the timepoint to the cluster level, thus preventing any conclusion to be made about the onset or offset of effects (e.g., differences across conditions or groups). Here, we introduce a *model-based* approch for analysing M/EEG timeseries such as ERPs or timecourses of decoding performance and their differences across conditions or groups. This approach relies on Bayesian generalised additive multilevel models, which output the posterior probabilility of the effect being above 0 (or above chance) at every timestep, while naturally taking into account the temporal dependencies and between-subject variability present in such data. Using both simulation and actual EEG data, we show that the proposed approach largely outperforms conventional methods in determining both the onset and offset of M/EEG effects (e.g., ERPs difference, decoding performance), producting more precise and more reliable estimates. We provide an R package implementing the approach and illustrate how to integrate it into M/EEG statistical pipelines in MNE-Python."
    )

available:::namr(
    title = "Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models",
    verb = TRUE
    )

# to-do adding some code here...
```
